Hierarchy Builder: algebraic~hierarchies made easy in Coq with Elpi

### 2020-04-25 hb.tex
typesetting + ref to appendixes

The first version of our hierarchy (that we name V1) features only two
structures: \coq{Monoid} and \coq{Ring}.
%\marginpar{remove hb.end for \mixins{}, use next release of coq elpi}
\begin{coqcode}
#\hbmixin{}# Record Monoid_of_Type A := {                       #\label{demo:mixin:mot}#
  zero : A;
  add : A -> A -> A;
  addrA : associative add;             (* `add` is associative.                   *)
  add0r : left_id zero add;            (* `zero` is the left and right neutral    *)
  addr0 : right_id zero add;           (*   element with respect to `add`.        *)
}.                                                             #\label{demo:mixin:mot:end}#
#\hbstructure{}# Definition Monoid := { A of Monoid_of_Type A }.        #\label{demo:structure:monoid}#

@@

Lemma addrC : forall R : Ring.type, commutative (@add R).                       #\label{demo:theory:state:addrC}#
Proof. (* Proof by Hankel 1867, omitted for brevity *) Qed.

@@

\begin{coqcode}
#\hbfactory{}# Record Ring_of_Monoid A of Monoid A := {
  ... (* unchanged, see page 2 line 13 *) ...
}.
#\hbbuilders{}# Context A (f : Ring_of_Monoid A).                          #\label{demo3:variable:f}#

@@

\subsection{Multiple inheritance}
In order to introduce multiple inheritance, we extend the hierarchy described in
section~\ref{subsec:v3} with the structure of semirings%, which is another
%intermediate structure between monoids and rings
as depicted in figure \ref{fig:puzzle_v4}. Semirings introduce the binary
multiplication operator \coq{mul} and its identity element \coq{one}.

@@

\section{Coq reference}

\begin{coqcode}
Section OperationProperties.
Variable T : Type.
Variable e : T.
Variable inv : T -> T.
Variable op : T -> T -> T.
Variable add : T -> T -> T.

@@

Definition left_inverse := forall x, op (inv x) x = e.

@@

Definition left_distributive  := forall x y z, op (add x y) z = add (op x z) (op y z).
Definition right_distributive := forall x y z, op x (add y z) = add (op x y) (op x z).
\end{coqcode}

\section{Proof of addrC}
\begin{coqcode}
Lemma addrC {R : Ring.type} : commutative (@add R).
Proof.
have innerC (a b : R) : (a + b) + (a + b) = (a + a) + (b + b).
  by rewrite -[a+b]mul1r -mulrDl mulrDr !mulrDl !mul1r.
have addKl (a b c : R) : a + b = a + c -> b = c.
  apply: can_inj (add a) (add (-a)) _ _ _.
  by move=> x; rewrite addrA addNr add0r.
have addKr (a b c : R) : b + a = c + a -> b = c.
  apply: can_inj (add ^~ a) (add ^~ (-a)) _ _ _.
  by move=> x; rewrite /= -addrA addrN addr0.
move=> x y; apply: addKl (x) _ _ _; apply: addKr (y) _ _ _.
by rewrite -!addrA [in RHS]addrA innerC !addrA.
Qed.
\end{coqcode}

\section{Full V4 code}
\begin{coqcode}
From Coq Require Import ssreflect ssrfun ZArith.
From HB Require Import structures.

@@

HB.mixin Record Monoid_of_Type A := {
  zero : A;
  add : A -> A -> A;
  addrA : associative add;
  add0r : left_id zero add;
  addr0 : right_id zero add;
}.
HB.structure Definition Monoid := { A of Monoid_of_Type A }.
Notation "0" := zero : hb_scope.
Infix "+" := (@add _) : hb_scope.

HB.mixin Record AbelianGroup_of_Monoid A of Monoid A := {
  opp : A -> A;
  addrC : commutative (add : A -> A -> A);
  addNr : left_inverse zero opp add;
}.
HB.structure Definition AbelianGroup := { A of Monoid A & AbelianGroup_of_Monoid A }.
Notation "- x" := (@opp _ x) : hb_scope.
Notation "x - y" := (x + - y) : hb_scope.

HB.mixin Record SemiRing_of_Monoid A of Monoid A := {
  one : A;
  mul : A -> A -> A;
  mulrA : associative mul;
  mul1r : left_id one mul;
  mulr1 : right_id one mul;
  mulrDl : left_distributive mul add;
  mulrDr : right_distributive mul add;
  mul0r : left_zero zero mul;
  mulr0 : right_zero zero mul;
}.
HB.structure Definition SemiRing := { A of Monoid A & SemiRing_of_Monoid A }.
Notation "1" := one : hb_scope.
Infix "*" := (@mul _) : hb_scope.

@@

HB.factory Record Ring_of_AbelianGroup A of AbelianGroup A := {
  one : A;
  mul : A -> A -> A;
  mulrA : associative mul;
  mul1r : left_id one mul;
  mulr1 : right_id one mul;
  mulrDl : left_distributive mul add;
  mulrDr : right_distributive mul add;
}.

HB.builders Context (A : Type) (f : Ring_of_AbelianGroup A).

@@

HB.factory Record Ring_of_Monoid A of Monoid A := {
  one : A;
  opp : A -> A;
  mul : A -> A -> A;
  addNr : left_inverse zero opp add;
  addrN : right_inverse zero opp add;
  mulrA : associative mul;
  mul1r : left_id one mul;
  mulr1 : right_id one mul;
  mulrDl : left_distributive mul add;
  mulrDr : right_distributive mul add;
}.

@@

  Definition to_AbelianGroup_of_Monoid := AbelianGroup_of_Monoid.Build A opp_f addrC addNr_f.
  HB.instance A to_AbelianGroup_of_Monoid.

  Definition to_Ring_of_AbelianGroup := Ring_of_AbelianGroup.Build A one_f mul_f
    mulrA_f mul1r_f mulr1_f mulrDl_f mulrDr_f.
  HB.instance A to_Ring_of_AbelianGroup.
HB.end.
\end{coqcode}

>>>>

The first version of our hierarchy (that we name V1) features only two
structures: \coq{Monoid} and \coq{Ring}. We use notations from Appendix~\ref{sec:coq-reference} to define them.
%\marginpar{remove hb.end for \mixins{}, use next release of coq elpi}
\begin{coqcode}
#\hbmixin{}# Record Monoid_of_Type A := {                       #\label{demo:mixin:mot}#
  zero : A;
  add : A -> A -> A;
  addrA : associative add;             (* `add` is associative.                   *)
  add0r : left_id zero add;            (* `zero` is the left and right neutral    *)
  addr0 : right_id zero add;           (*   element with respect to `add`.        *)
}.                                                             #\label{demo:mixin:mot:end}#
#\hbstructure{}# Definition Monoid := { A of Monoid_of_Type A }.        #\label{demo:structure:monoid}#

@@

Lemma addrC : forall R : Ring.type, commutative (@add R).                       #\label{demo:theory:state:addrC}#
Proof. (* Proof by Hankel 1867, in Appendix B *) Qed.

@@

\begin{coqcode}
#\hbfactory{}# Record Ring_of_Monoid A of Monoid A := { ... (* unchanged *) ...}.

#\hbbuilders{}# Context A (f : Ring_of_Monoid A).                          #\label{demo3:variable:f}#

@@

\subsection{Multiple inheritance}
In order to introduce multiple inheritance, we extend the hierarchy described in
section~\ref{subsec:v3} with the structure of semirings%, which is another
%intermediate structure between monoids and rings
as depicted in figure \ref{fig:puzzle_v4}. Semirings introduce the binary
multiplication operator \coq{mul} and its identity element \coq{one}.
The full code of this V4 is available in Appendix~\ref{sec:full-v4-code}.

@@

\section{Coq reference}
\label{sec:coq-reference}
\begin{coqcode}
Section OperationProperties.

Variables (T : Type) (e : T) (inv : T -> T) (op : T -> T -> T) (add : T -> T -> T).

@@

Definition left_inverse := forall x, op (inv x) x = e.
Definition right_inverse := forall x, op x (inv x) = e.

@@

Definition left_distributive  := forall x y z, op (add x y) z = add (op x z) (op y z).
Definition right_distributive := forall x y z, op x (add y z) = add (op x y) (op x z).

End OperationProperties.
\end{coqcode}

\section{Proof of addrC}
\label{sec:proof-addrc}
\begin{coqcode}
Lemma addrC {R : Ring.type} : commutative (@add R).
Proof.
have innerC (a b : R) : (a + b) + (a + b) = (a + a) + (b + b).
  by rewrite -[a+b]mul1r -mulrDl mulrDr !mulrDl !mul1r.
have addKl (a b c : R) : a + b = a + c -> b = c.
  apply: can_inj (add a) (add (-a)) _ _ _.
  by move=> x; rewrite addrA addNr add0r.
have addKr (a b c : R) : b + a = c + a -> b = c.
  apply: can_inj (add ^~ a) (add ^~ (-a)) _ _ _.
  by move=> x; rewrite /= -addrA addrN addr0.
move=> x y; apply: addKl (x) _ _ _; apply: addKr (y) _ _ _.
by rewrite -!addrA [in RHS]addrA innerC !addrA.
Qed.
\end{coqcode}

\section{Full V4 code}
\label{sec:full-v4-code}
\begin{coqcode}
From Coq Require Import ssreflect ssrfun ZArith.
From HB Require Import structures.

@@

(* Bottom mixin in Fig. 2. *)
HB.mixin Record Monoid_of_Type A := {
  zero : A;
  add : A -> A -> A;
  addrA : associative add;
  add0r : left_id zero add;
  addr0 : right_id zero add;
}.
HB.structure Definition Monoid := { A of Monoid_of_Type A }.
Notation "0" := zero : hb_scope.
Infix "+" := (@add _) : hb_scope.

(* Bottom right mixin in Fig. 2. *)
HB.mixin Record AbelianGroup_of_Monoid A of Monoid A := {
  opp : A -> A;
  addrC : commutative (add : A -> A -> A);
  addNr : left_inverse zero opp add;
}.
HB.structure Definition AbelianGroup := { A of Monoid A & AbelianGroup_of_Monoid A }.
Notation "- x" := (@opp _ x) : hb_scope.
Notation "x - y" := (x + - y) : hb_scope.

(* Bottom left mixin in Fig. 2. *)
HB.mixin Record SemiRing_of_Monoid A of Monoid A := {
  one : A;
  mul : A -> A -> A;
  mulrA : associative mul;
  mul1r : left_id one mul;
  mulr1 : right_id one mul;
  mulrDl : left_distributive mul add;
  mulrDr : right_distributive mul add;
  mul0r : left_zero zero mul;
  mulr0 : right_zero zero mul;
}.
HB.structure Definition SemiRing := { A of Monoid A & SemiRing_of_Monoid A }.
Notation "1" := one : hb_scope.
Infix "*" := (@mul _) : hb_scope.

@@

(* Top right factory in Fig. 2. *)
HB.factory Record Ring_of_AbelianGroup A of AbelianGroup A := {
  one : A;
  mul : A -> A -> A;
  mulrA : associative mul;
  mul1r : left_id one mul;
  mulr1 : right_id one mul;
  mulrDl : left_distributive mul add;
  mulrDr : right_distributive mul add;
}.

(* Builder arrow from top right to bottom left *)
HB.builders Context (A : Type) (f : Ring_of_AbelianGroup A).

@@

(* Top left factory in Fig. 2. *)
(* It is an exact copy of the bottom right mixin. *)
HB.factory Definition Ring_of_SemiRing A of SemiRing A := AbelianGroup_of_Monoid A.
(* The corresponding builder is the identity. *)
HB.builders Context (A : Type) (f : Ring_of_SemiRing A).
  Definition to_AbelianGroup_of_Monoid : AbelianGroup_of_Monoid A := f.
  HB.instance A to_AbelianGroup_of_Monoid.
HB.end.

(* Right-most factory in Fig. 2. *)
HB.factory Record Ring_of_Monoid A of Monoid A := {
  one : A;
  opp : A -> A;
  mul : A -> A -> A;
  addNr : left_inverse zero opp add;
  addrN : right_inverse zero opp add;
  mulrA : associative mul;
  mul1r : left_id one mul;
  mulr1 : right_id one mul;
  mulrDl : left_distributive mul add;
  mulrDr : right_distributive mul add;
}.

@@

  (* Builder to the bottom right mixin. *)
  Definition to_AbelianGroup_of_Monoid := AbelianGroup_of_Monoid.Build A opp_f addrC addNr_f.
  HB.instance A to_AbelianGroup_of_Monoid.

  (* Builder to the top right factory, which is compiled to the bottom left mixin. *)
  Definition to_Ring_of_AbelianGroup := Ring_of_AbelianGroup.Build A one_f mul_f
    mulrA_f mul1r_f mulr1_f mulrDl_f mulrDr_f.
  HB.instance A to_Ring_of_AbelianGroup.

HB.end.
\end{coqcode}
